Evolution of ORM Systems: Formal Foundations

### 2013-01-23 doc/description.tex
formatting

\documentclass[11pt]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{semantic}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{algpseudocode}

@@

\section{ORM Software}
\label{sec:evoIntro}
\begin{figure}
\centering
	\includegraphics[scale=0.3]{./images/system}
	\caption{The model of one generation of an ORM system consists of application persistent layer and relational database for storing instances.}
\label{fig:appStructure}
\end{figure}
A software implemented using an object-oriented language, which uses relational database as a data storage (as illustrated in the Figure \ref{fig:appStructure}) is object of interest of this paper. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM (we will use $\rho$ as a symbol for ORM), therefore we define a software as a triple:
\begin{align*}
	\mathbf{software} = < Application, Database, \rho > \\
	sw : Applications \times Databases \times ORM \rightarrow Software \\ \\
	\mathbf{application} : Software \rightarrow Application \\
	application(sw(a, d, \rho)) = a \\ \\
	\mathbf{database} : Software \rightarrow Database \\
	database(sw(a, d, \rho)) = d \\
	a \lhd Application, d \lhd Database
\end{align*}
The software definition above introduce the first type used in this paper. The notation $a \lhd Application$ indicates the variable $a$ has the type $Application$.

The software has to be in consistent state so its users could have benefit from its usage. The state of the software is defined to communicate the sate of the software to its user. The components of a software (application and database) are transformed likewise, thus the state is defined for them too.
\begin{align*}
 	\mathbf{state} = Consistent\; s \: | \: Inconsistent \; s,  \\	s \lhd State \vee s \lh Application \vee s \lhd Database
\end{align*}
Before the definition of the software consistency is provided, the notation of function transcription has to be introduced. Usually the definition of a function consists of two parts - an effect of the function and conditions when the effect occurs We decide to use following transcription, to simplify the notation and to improve understanding of functions:
\begin{align*}
f(x) =   \inference{conditions}{effect}
\end{align*}

The function querying the state (it means consistency) of a software is defined as:
\begin{align*}
 	\mathbf{consistent} : State \rightarrow Boolean \\
 	consistent(s) = \begin{cases}
 		\inference{s = Consistent(s)}{true}\\\\
 		\inference{s = Inconsistent(s)}{false}\\
 \end{cases}\\
 s \lhd State
\end{align*}
The consistency of application and database is discussed later in this paper. The definition of software state only is provided:
\begin{align*}
	\mathbf{state} : Software \rightarrow State \\
	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \\ \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases} \\ 
 	s \lhd State
\end{align*}

@@

The evolution of the whole software has to be processed on the both software components - application and database. It means there is a set of application transformations called $AppT$ and a set of database transformations called $DbT$. Because there are needed different parameters for application and database evolution the evolution of database can not be derived from application evolution only. Thus the set of transformation called $E$ is defined to evolve the whole software together with mappings $\Psi$ between $E$ and $AppT$ and $\Phi$ between $E$ and $DbT$ is defined:
\begin{align*}
\Psi : E \rightarrow AppT \\
\Phi : E \rightarrow DbT
\end{align*}
The set $E$ containing the evolution transformations of a software mets our first goal - a single source of data evolution for the whole software.

The application of a transformation on a software state (evolution) is marked by the $\oplus$ operator:
$$\oplus : State \times E \rightarrow State $$
and similarly the $\circ$ and the $\bullet$ operators for evolving application and database:
\begin{align*}
\circ : State \times AppT \rightarrow State \\
 \bullet : State \times DbT \rightarrow State
\end{align*}
The definition of the evolving operator is inspired by the monads (as used e.g. in Haskell programming language) so the state of the software can be passed from one evolution step to another, which helps us to concatenate evolution transformations.
\begin{align*}
 state(s) \oplus t = \begin{cases}
 	\inference{! consistent(state(s))}{u} \\ \\
 	\inference{consistent(state(s))}{state(sw(\Psi(s, t), \Phi(s,t))} 
\end{cases}\\ 
s \lhd State, t \in E
\end{align*}
\begin{align*}
 state(a) \circ e = \begin{cases}
 	\inference{!\;consistent(state(a))}{a} \\ \\
 	\inference{consistent(state(a))}{e(a)} 
\end{cases}\\ 
a \lhd Application, e \in AppT
\end{align*}
\begin{align*}
 state(d) \bullet e = \begin{cases}
 	\inference{!\;consistent(state(d))}{d} \\ \\
 	\inference{consistent(state(d))}{e(d)} 
\end{cases}\\ 
d \lhd Application, e \in DbT
\end{align*}

@@

The definitions of all parts assume a set $Label$ exists which contains all possible identifiers and each modeled entity is unambiguously identified in the model. Next there is a set $Values$ containing all possible values, which can be used in the model. The notation: $Type*$ is used to define an ordered set of values of concrete type (e.g. $C \lhd Class*$ means c is an ordered set of classes).

A lot of introduced transformations can be applied on an ordered set. The application proceeds on set consists in sequential application on set elements:
\begin{align*}
	f(x, y) = \begin{cases}
 		\inference{x = \varnothing}{y}\\
 		\inference{|x| > 1}{f(head(x), y) \cup f(tail(x), y}\\\\
 		\inference{|x| = 1}{effect}
 		\inference{otherwise}{Inconsistent y}
 \end{cases}\\
 y \lhd Software \vee y \lhd Application \vee y \lhd Database
\end{align*}
where $x$ is a set of parameters. The effect of a transformation is provided only in next sections as the repetition of sequential application will be redundant. The $otherwise$ branch is consider implicit for all introduced transformations and thus it is not part of transformation definitions too.

\subsection{Application Model}
\label{sec:appModel}
The application model represents an application persistent layer created using object oriented language and functions which can alter the model. 
\subsubsection{Application Structure}
Each instance in an application is defined according to some class - the class defines the structure of its instances and relationships between them an the rest of the application. The instances are stored in the relational database in our case, therefore we define only the structure of classes in our model. 
\begin{align*}
  \mathbf{Application} = < Class* > \\ 
  \mathbf{app} : Class* \rightarrow  Application \\ \\
  \mathbf{classes} : Application \rightarrow Class* \\
  classes(app(cs)) = cs  \\\\
  \mathbf{class} : Label \times Application \rightarrow Class   \\ 
  class(l, a) = c, c \in classes(a) \wedge name(c) = l \\ \\
  l \in Label,a \lhd Applications, cs = Class* 
\end{align*}

\paragraph{Class} Class represents a basic structural unit in the application model. It has a unique name, one or more properties and it can be associated to other classes in the application. 
\begin{align*}
	\mathbf{Class} = <label, Property*, Association*> \\
	\mathbf{cl} : Label \times Property* \times Association* \rightarrow Class \\\\
	\mathbf{name} : Class \rightarrow Label \\
	name(cl(l, props, assocs)) = l \\ \\
	\mathbf{properties} : Class \rightarrow Property* \\
	properties(cl(l, props, assocs)) = props \\ \\
	\mathbf{associations} : Class \rightarrow Association* \\	associations(cl(l, props, assocs)) = assocs \\ 
	l \in Label, props \lhd Property*, assocs \lhd Association*
\end{align*}
Three more function are defined to obtain information about a class - function $primitives$ returns all properties of the class which represent a single value, whereas function $collections$ returns all properties representing collection of values. The last function called $associated$ determine if a class is associated by any other class in an application.
\begin{align*}
	\mathbf{primitives} : Class \rightarrow Property* \\	primitives(cl(l, props, assocs)) = props', \\ props' \subset props,  \forall p \in props' : cardinality(p) \leq 1 \\ \\
	\mathbf{collections} : Class \rightarrow Property* \\	collections(cl(l, props, assocs)) = props', \\ props' \subset props, \forall p \in props' : cardinality(p) > 1 \\ \\
	\mathbf{associated} :  Label \times Application \rightarrow Boolean \\
  	associated(l,a) = \begin{cases}
  		\inference{\exists c \in classes(a) \wedge \exists e \in associations(c) : \\ reference(e) = l
  	}{ true } \\\\
 		false
 	\end{cases} \\ \\
	l \in Label, props = Property*, assocs = Association*,a \lhd Applications
\end{align*}

\paragraph{Property} Property represents a feature of a class which is represented as a primitive type. The property can be mandatory, can have a default value and according to its cardinality it can represent a single value or a collection of values. The $owningClass$ function returns the class in the model which owns the property.
\begin{align*}
	\mathbf{Property} = <label, AppType, DefaultValue, Cardinality, Mandatory> \\
	\mathbf{prop} : Label \times AppType \times String \times \mathbb{N^{*}} \times Boolean \rightarrow Property \\ \\
	\mathbf{name} : Property \rightarrow Label \\
	name(prop(l, t, val, n, m)) = l \\ \\
	\mathbf{type} : Property \times AppType \\
	type(prop(l, t, val, n, m)) = t \\ \\
	\mathbf{cardinality} : Property \rightarrow \mathbb{N^{*}} \\
	cardinality(prop(l, t, val, n, m)) = n \\ \\
	\mathbf{mandatory} : Property \rightarrow Boolean \\
	mandatory(prop(l, t, val, n, m)) = m  \\ \\
	\mathbf{owningClass} : Property \times Application \rightarrow Class  \\ 	
	owningClass(p, a) = \inference{\exists c \in classes(a) \wedge p \in properties(c)}{c}  \\ \\
	l \in Label, t \lhd AppType, val \in Values, n \in \mathbb{N^{*}}, m \in Boolean 
\end{align*}

\paragraph {Association} Associtation represents a connection between two classes. It has a unique name and the reference is represented by the label of referenced class. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of both association ends.
\begin{align*}
	\mathbf{Association} = <label, classRef, startCardinality, endCardinality> \\
	\mathbf{assoc} : Label \times Label \times \mathbb{N} \times \mathbb{N} \rightarrow Association\\ \\
	\mathbf{name} : Association \rightarrow Label \\
	name(assoc(l, ref, n_1, n_2)) = l\\ \\
	\mathbf{reference} : Association \rightarrow Label \\
	reference(assoc(l, ref, n_1, n_2)) = ref\\ \\
	\mathbf{startCardinality} : Association \rightarrow \mathbb{N} \\
	startCardinality(assoc(l, ref, n_1, n_2)) = n_1\\ \\
	\mathbf{endCardinality} : Association \rightarrow \mathbb{N} \\
	endCardinality(assoc(l, ref, n_1, n_2)) = n_2 \\ \\
	\mathbf{owningClass} : Association \times Application \rightarrow Class  \\
	owningClass(as, a) = \inference{\exists c \in classes(a) \wedge as \in associations(c)}{c}  \\ \\	
	l, ref \in Label,  n_1, n_2 \in \mathbb{N}, as \lhd Association
\end{align*}

@@

\subsubsection{Application consistency}
The application consistency is based on the structural consistency of its model. Violation of structural consistency are name collisions and non-existing references in associations. The state of an application is:
\begin{align*}
	state(a) = \begin{cases}
	\inference{\forall c_1, c_2 \in classes(a) : name(c_1) \neq name(c_2) \\ \wedge \forall p_1, p_2 \in properties(c_1) : name(p_1) \neq name(p_2) \\ \wedge \forall a_1, a_2 \in associations(c_1) : name(a_1) \neq name (a_2) \\ \wedge \forall a \in associations(c_1) : \exists c \in classes(a) \wedge name(c) = reference(a) }{Consistent \; a}\\\\
 	Inconsistent \; a
 	\end{cases}\\
	a \lhd Application
\end{align*}

\subsection{Database Model}
\label{sec:dbModel}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents stored instances. The database is defined as a tuple:
\begin{align*}
	\mathbf{Database} = < Table*, Row*> \\
	\mathbf{db} : Tables \times Rows \rightarrow Database \\ \\
	\mathbf{tables} : Database \rightarrow Tables \\
	tables(db(tabs, rs)) = tabs \\ \\
	\mathbf{rows} : Database \rightarrow Rows \\
	rows(db(tabs, rs)) = rs \\ \\
	tabs = Table*, rs = Row*, l \in Label,d \lhd Database
\end{align*}
\paragraph{Table} Table represents a basic concept of database schema. It has a unique name, one or more columns and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
\begin{align*}
	\mathbf{Table} = <label, primaryKey, Column*, ForeignKey*>\\
	\mathbf{tab} 	: Label \times PrimaryKeys \times Columns \times ForeignKeys \rightarrow Table \\ \\
	\mathbf{name} : Table \rightarrow Label  \\
	name(tab(l, pk, cols, fks)) = l \\ \\
	\mathbf{primaryKey} : Table \rightarrow PrimaryKeys  \\
	primaryKey(tab(l, pk, cols, fks)) = pk \\ \\
	\mathbf{columns} : Table \rightarrow Columns  \\
	columns(tab(l, pk, cols, fks)) = cols \\ \\
	\mathbf{foreignKeys} : Table \rightarrow ForeignKeys  \\
	foreignKeys(tab(l, pk, cols, fks)) = fks \\ \\
l \in Label, pk \lhd PrimaryKey, cols = Column*, fks = ForeignKey*
\end{align*}

\paragraph{Column} Column defines data values and types which can be part of a table record.
\begin{align*}
	\mathbf{Column} = <label, DbType, DefaultValue, Constraint*> \\
	\mathbf{col} : Label \times DbType \times String \times Constraints \rightarrow Column \\ \\
	\mathbf{name} : Column \rightarrow Label  \\
	name(col(l, t, val, cons)) = l  \\ \\
	\mathbf{type} : Column \rightarrow DbType  \\
	type(col(l, t, val, cons)) = t  \\ \\
	\mathbf{constraints} : Column \rightarrow Constraints  \\
	constraints(col(l, t, val, cons)) = cons  \\ \\
	\mathbf{owningTable} : Column \times Database \rightarrow Table  \\
	owningTable(c, d) = \inference{\exists tab \in tables(d) \wedge c \in columns(tab)}{tab} \\ \\
l \in Label, t \lhd DbType, val \in Values,\\ cons \lhd Constraint*, c \lhd Column,d \lhd Database
\end{align*}

\paragraph{Foreign key} Foreign key is a reference to another table's primary key, it has an unique name and it can be constrained. The value of a foreign key is $\varnothing$ or a non-zero natural number.
\begin{align*}
	\mathbf{ForeignKey} = <label, tableRef, Constraint*> \\ 
	\mathbf{fk} : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
	\mathbf{name} : ForeignKey \rightarrow Label \\
	name(fk(l, tRef, cons)) = lab  \\ \\
	\mathbf{reference} : ForeignKey \rightarrow Label  \\
	reference(fk(l, tRef, cons)) = tRef  \\ \\
	\mathbf{constaints} : ForeignKey \rightarrow Constraint*  \\
	constaints(fk(l, tRef, cons)) = cons  \\ \\
	\mathbf{owningTable} : ForeignKey \times Database \rightarrow Table  \\
	owningTable(fk, d) = \inference{\exists tab \in tables(d) \wedge fk \in foreignKeys(tab)}{tab}\\\\
	l, tRef \in Label, cons \lhd Constraint*,d \lhd Database, fk \lhd ForeignKey
\end{align*}

@@

\paragraph{Mapping of Classes} The class is mapped into a table with  automatically created primary key. The primary key is always created with name Id and of the $\mathbb{N^{*}}$ type. The properties and associations has to be mapped separately.
\begin{align*}
	\mathbf{\rho_{c}} : Class*  \rightarrow Table* \\
	\rho_{c}(c) = \begin{cases}
		%\inference{c = \varnothing}{\varnothing}\\\\
		%\inference{|c| > 1}{\rho_c(head(c)) \cup \rho_c(tail(c))} \\ \\
		\inference{|c| = 1}{table(name(c), primaryKey("Id"), \varnothing, \varnothing)}  \end{cases}   \\ 	
 c \lhd Class
\end{align*}

@@

%\paragraph{Get Property or Association} Functions $findProperty$ and $findAssociation$ returns a property (association) if it exists in given context - application and class of provided name.
%\begin{align*}
%	\mathbf{findProperty} : Label \times Label \times Application \rightarrow Property \\
%	findProperty(l_c, l_p, a) = \inference{\exists c \in classes(a): name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_p}{p} \\
%	l_c, l_p \in Label, a \lhd Application
%\end{align*}
%\begin{align*}
%	\mathbf{findAssociation} : Label \times Label \times Application \rightarrow Association \\
%	findAssociation(l_c, l_a, a) = \inference{\exists c \in classes(a): name(c) = l_c \\ \wedge \exists o \in associations(c) : name(o) = l_a}{o} \\
%	l_c, l_a \in Label, a \lhd Application
%\end{align*}

\paragraph{Add Property} The name collision only has to be verified, before adding a property into an application structure:
\begin{align*}
 	\mathbf{addProp} : Label \times Property \times Application \rightarrow Application \\
	addProp(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(c) : name(q) \neq name(p) \\ \wedge \forall e \in associations(c) : name(e) \neq name(p)}{properties(c) \cup p } \\
	l \in Label, p \lhd Property, a \lhd Applications
\end{align*}

>>>>

\documentclass[11pt]{article}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{semantic}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{algpseudocode}

@@

\section{ORM Software}
\label{sec:evoIntro}
\begin{figure}
\centering
	\includegraphics[scale=0.3]{./images/system}
	\caption{The model of one generation of an ORM system consists of application persistent layer and relational database for storing instances.}
\label{fig:appStructure}
\end{figure}
A software implemented using an object-oriented language, which uses relational database as a data storage (as illustrated in the Figure \ref{fig:appStructure}) is object of interest of this paper. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM (we will use $\rho$ as a symbol for ORM), therefore we define a software as a triple:
\begin{align*}
& \mathbf{software} = < Application, Database, \rho >\\
& \mathbf{sw} : Applications \times Databases \times ORM \rightarrow  Software \\ \\
& \mathbf{application} : Software \rightarrow Application \\
& application(sw(a, d, \rho)) = a \\ \\
& \mathbf{database} : Software \rightarrow Database \\
& database(sw(a, d, \rho)) = d \\\\
& a : Application, d : Database
\end{align*}

The software has to be in consistent state so its users could have benefit from its usage. The state of the software is defined to communicate the sate of the software to its user. The components of a software (application and database) are transformed likewise, thus the state is defined for them too.
\begin{align*}
& 	\mathbf{state} = Consistent\; s \: | \: Inconsistent \; s,  \\\\
& 	s : State \vee s \lh Application \vee s : Database
\end{align*}
Before the definition of the software consistency is provided, the notation of function transcription has to be introduced. Usually the definition of a function consists of two parts - an effect of the function and conditions when the effect occurs We decide to use following transcription, to simplify the notation and to improve understanding of functions:
\begin{align*}
f(x) =   \inference{conditions}{effect}
\end{align*}

The function querying the state (it means consistency) of a software is defined as:
\begin{align*}
& 	\mathbf{consistent} : State \rightarrow Boolean \\
& 	consistent(s) = \begin{cases}
 		\inference{s = Consistent(s)}{true}\\\\
 		\inference{s = Inconsistent(s)}{false}\\
 \end{cases}\\
& s : State
\end{align*}
The consistency of application and database is discussed later in this paper. The definition of software state only is provided:
\begin{align*}
&	\mathbf{state} : Software \rightarrow State \\
&	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \\ \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases} \\ 
& 	s : State
\end{align*}

@@

The evolution of the whole software has to be processed on the both software components - application and database. It means there is a set of application transformations called $AppT$ and a set of database transformations called $DbT$. Because there are needed different parameters for application and database evolution the evolution of database can not be derived from application evolution only. Thus the set of transformation called $E$ is defined to evolve the whole software together with mappings $\Psi$ between $E$ and $AppT$ and $\Phi$ between $E$ and $DbT$ is defined:
\begin{align*}
&	\Psi : E \rightarrow AppT \\
&	\Phi : E \rightarrow DbT
\end{align*}
The set $E$ containing the evolution transformations of a software mets our first goal - a single source of data evolution for the whole software.

The application of a transformation on a software state (evolution) is marked by the $\oplus$ operator:
\begin{align*}
& \oplus : State \times E \rightarrow State
\end{align*}
and similarly the $\circ$ and the $\bullet$ operators for evolving application and database:
\begin{align*}
& \circ : State \times AppT \rightarrow State \\
& \bullet : State \times DbT \rightarrow State
\end{align*}
The definition of the evolving operator is inspired by the monads (as used e.g. in Haskell programming language) so the state of the software can be passed from one evolution step to another, which helps us to concatenate evolution transformations.
\begin{align*}
& state(s) \oplus t = \begin{cases}
 	\inference{! consistent(state(s))}{u} \\ \\
 	\inference{consistent(state(s))}{state(sw(\Psi(s, t), \Phi(s,t))} 
\end{cases}\\ 
& s : State, t \in E
\end{align*}
\begin{align*}
& state(a) \circ e = \begin{cases}
 	\inference{!\;consistent(state(a))}{a} \\ \\
 	\inference{consistent(state(a))}{e(a)} 
\end{cases}\\ 
& a : Application, e \in AppT
\end{align*}
\begin{align*}
&  state(d) \bullet e = \begin{cases}
 	\inference{!\;consistent(state(d))}{d} \\ \\
 	\inference{consistent(state(d))}{e(d)} 
\end{cases}\\ 
& d : Application, e \in DbT
\end{align*}

@@

The definitions of all parts assume a set $Label$ exists which contains all possible identifiers and each modeled entity is unambiguously identified in the model. Next there is a set $Values$ containing all possible values, which can be used in the model. The notation: $Type*$ is used to define an ordered set of values of concrete type (e.g. $C : Class*$ means c is an ordered set of classes).

A lot of introduced transformations can be applied on an ordered set. The application proceeds on set consists in sequential application on set elements:
\begin{align*}
& 	f(x, y) = \begin{cases}
 		\inference{x = \varnothing}{y}\\
 		\inference{|x| > 1}{f(head(x), y) \cup f(tail(x), y}\\\\
 		\inference{|x| = 1}{effect}\\\\
 		\inference{otherwise}{Inconsistent y}
 \end{cases}\\\\
&  y : Software \vee y : Application \vee y : Database
\end{align*}
where $x$ is a set of parameters. The effect of a transformation is provided only in next sections as the repetition of sequential application will be redundant. The $otherwise$ branch is consider implicit for all introduced transformations and thus it is not part of transformation definitions too.

\subsection{Application Model}
\label{sec:appModel}
The application model represents an application persistent layer created using object oriented language and functions which can alter the model. 
\subsubsection{Application Structure}
Each instance in an application is defined according to some class - the class defines the structure of its instances and relationships between them an the rest of the application. The instances are stored in the relational database in our case, therefore we define only the structure of classes in our model. 
\begin{align*}
& \mathbf{Application} = < Class* > \\ 
& \mathbf{app} : Class* \rightarrow  Application \\ \\
& \mathbf{classes} : Application \rightarrow Class* \\
&   classes(app(cs)) = cs  \\\\
& \mathbf{class} : Label \times Application \rightarrow Class   \\ 
&  class(l, a) = c, c \in classes(a) \wedge name(c) = l \\ \\
&  l \in Label,a : Applications, cs = Class* 
\end{align*}

\paragraph{Class} Class represents a basic structural unit in the application model. It has a unique name, one or more properties and it can be associated to other classes in the application. 
\begin{align*}
& 	\mathbf{Class} = <label, Property*, Association*> \\
& 	\mathbf{cl} : Label \times Property* \times Association* \rightarrow Class \\\\
& 	\mathbf{name} : Class \rightarrow Label \\
& 	name(cl(l, props, assocs)) = l \\ \\
& 	\mathbf{properties} : Class \rightarrow Property* \\
& 	properties(cl(l, props, assocs)) = props \\ \\
& 	\mathbf{associations} : Class \rightarrow Association* \\ 
& 	associations(cl(l, props, assocs)) = assocs \\ 
& 	l \in Label, props : Property*, assocs : Association*
\end{align*}
Three more function are defined to obtain information about a class - function $primitives$ returns all properties of the class which represent a single value, whereas function $collections$ returns all properties representing collection of values. The last function called $associated$ determine if a class is associated by any other class in an application.
\begin{align*}
	\mathbf{primitives} : Class \rightarrow Property* \\	primitives(cl(l, props, assocs)) = props', \\ props' \subset props,  \forall p \in props' : cardinality(p) \leq 1 \\ \\
	\mathbf{collections} : Class \rightarrow Property* \\	collections(cl(l, props, assocs)) = props', \\ props' \subset props, \forall p \in props' : cardinality(p) > 1 \\ \\
	\mathbf{associated} :  Label \times Application \rightarrow Boolean \\
  	associated(l,a) = \begin{cases}
  		\inference{\exists c \in classes(a) \wedge \exists e \in associations(c) : \\ reference(e) = l
  	}{ true } \\\\
 		false
 	\end{cases} \\ \\
	l \in Label, props = Property*, assocs = Association*,a : Applications
\end{align*}

\paragraph{Property} Property represents a feature of a class which is represented as a primitive type. The property can be mandatory, can have a default value and according to its cardinality it can represent a single value or a collection of values. The $owningClass$ function returns the class in the model which owns the property.
\begin{align*}
	\mathbf{Property} = <label, AppType, DefaultValue, Cardinality, Mandatory> \\
	\mathbf{prop} : Label \times AppType \times String \times \mathbb{N^{*}} \times Boolean \rightarrow Property \\ \\
	\mathbf{name} : Property \rightarrow Label \\
	name(prop(l, t, val, n, m)) = l \\ \\
	\mathbf{type} : Property \times AppType \\
	type(prop(l, t, val, n, m)) = t \\ \\
	\mathbf{cardinality} : Property \rightarrow \mathbb{N^{*}} \\
	cardinality(prop(l, t, val, n, m)) = n \\ \\
	\mathbf{mandatory} : Property \rightarrow Boolean \\
	mandatory(prop(l, t, val, n, m)) = m  \\ \\
	\mathbf{owningClass} : Property \times Application \rightarrow Class  \\ 	
	owningClass(p, a) = \inference{\exists c \in classes(a) \wedge p \in properties(c)}{c}  \\ \\
	l \in Label, t : AppType, val \in Values, n \in \mathbb{N^{*}}, m \in Boolean 
\end{align*}

\paragraph {Association} Associtation represents a connection between two classes. It has a unique name and the reference is represented by the label of referenced class. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of both association ends.
\begin{align*}
	\mathbf{Association} = <label, classRef, startCardinality, endCardinality> \\
	\mathbf{assoc} : Label \times Label \times \mathbb{N} \times \mathbb{N} \rightarrow Association\\ \\
	\mathbf{name} : Association \rightarrow Label \\
	name(assoc(l, ref, n_1, n_2)) = l\\ \\
	\mathbf{reference} : Association \rightarrow Label \\
	reference(assoc(l, ref, n_1, n_2)) = ref\\ \\
	\mathbf{startCardinality} : Association \rightarrow \mathbb{N} \\
	startCardinality(assoc(l, ref, n_1, n_2)) = n_1\\ \\
	\mathbf{endCardinality} : Association \rightarrow \mathbb{N} \\
	endCardinality(assoc(l, ref, n_1, n_2)) = n_2 \\ \\
	\mathbf{owningClass} : Association \times Application \rightarrow Class  \\
	owningClass(as, a) = \inference{\exists c \in classes(a) \wedge as \in associations(c)}{c}  \\ \\	
	l, ref \in Label,  n_1, n_2 \in \mathbb{N}, as : Association
\end{align*}

@@

\subsubsection{Application consistency}
The application consistency is based on the structural consistency of its model. Violation of structural consistency are name collisions and non-existing references in associations. The state of an application is:
\begin{align*}
	state(a) = \begin{cases}
	\inference{\forall c_1, c_2 \in classes(a) : name(c_1) \neq name(c_2) \\ \wedge \forall p_1, p_2 \in properties(c_1) : name(p_1) \neq name(p_2) \\ \wedge \forall a_1, a_2 \in associations(c_1) : name(a_1) \neq name (a_2) \\ \wedge \forall a \in associations(c_1) : \exists c \in classes(a) \wedge name(c) = reference(a) }{Consistent \; a}\\\\
 	Inconsistent \; a
 	\end{cases}\\
	a : Application
\end{align*}

\subsection{Database Model}
\label{sec:dbModel}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents stored instances. The database is defined as a tuple:
\begin{align*}
	\mathbf{Database} = < Table*, Row*> \\
	\mathbf{db} : Tables \times Rows \rightarrow Database \\ \\
	\mathbf{tables} : Database \rightarrow Tables \\
	tables(db(tabs, rs)) = tabs \\ \\
	\mathbf{rows} : Database \rightarrow Rows \\
	rows(db(tabs, rs)) = rs \\ \\
	tabs = Table*, rs = Row*, l \in Label,d : Database
\end{align*}
\paragraph{Table} Table represents a basic concept of database schema. It has a unique name, one or more columns and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
\begin{align*}
	\mathbf{Table} = <label, primaryKey, Column*, ForeignKey*>\\
	\mathbf{tab} 	: Label \times PrimaryKeys \times Columns \times ForeignKeys \rightarrow Table \\ \\
	\mathbf{name} : Table \rightarrow Label  \\
	name(tab(l, pk, cols, fks)) = l \\ \\
	\mathbf{primaryKey} : Table \rightarrow PrimaryKeys  \\
	primaryKey(tab(l, pk, cols, fks)) = pk \\ \\
	\mathbf{columns} : Table \rightarrow Columns  \\
	columns(tab(l, pk, cols, fks)) = cols \\ \\
	\mathbf{foreignKeys} : Table \rightarrow ForeignKeys  \\
	foreignKeys(tab(l, pk, cols, fks)) = fks \\ \\
l \in Label, pk : PrimaryKey, cols = Column*, fks = ForeignKey*
\end{align*}

\paragraph{Column} Column defines data values and types which can be part of a table record.
\begin{align*}
	\mathbf{Column} = <label, DbType, DefaultValue, Constraint*> \\
	\mathbf{col} : Label \times DbType \times String \times Constraints \rightarrow Column \\ \\
	\mathbf{name} : Column \rightarrow Label  \\
	name(col(l, t, val, cons)) = l  \\ \\
	\mathbf{type} : Column \rightarrow DbType  \\
	type(col(l, t, val, cons)) = t  \\ \\
	\mathbf{constraints} : Column \rightarrow Constraints  \\
	constraints(col(l, t, val, cons)) = cons  \\ \\
	\mathbf{owningTable} : Column \times Database \rightarrow Table  \\
	owningTable(c, d) = \inference{\exists tab \in tables(d) \wedge c \in columns(tab)}{tab} \\ \\
l \in Label, t : DbType, val \in Values,\\ cons : Constraint*, c : Column,d : Database
\end{align*}

\paragraph{Foreign key} Foreign key is a reference to another table's primary key, it has an unique name and it can be constrained. The value of a foreign key is $\varnothing$ or a non-zero natural number.
\begin{align*}
	\mathbf{ForeignKey} = <label, tableRef, Constraint*> \\ 
	\mathbf{fk} : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
	\mathbf{name} : ForeignKey \rightarrow Label \\
	name(fk(l, tRef, cons)) = lab  \\ \\
	\mathbf{reference} : ForeignKey \rightarrow Label  \\
	reference(fk(l, tRef, cons)) = tRef  \\ \\
	\mathbf{constaints} : ForeignKey \rightarrow Constraint*  \\
	constaints(fk(l, tRef, cons)) = cons  \\ \\
	\mathbf{owningTable} : ForeignKey \times Database \rightarrow Table  \\
	owningTable(fk, d) = \inference{\exists tab \in tables(d) \wedge fk \in foreignKeys(tab)}{tab}\\\\
	l, tRef \in Label, cons : Constraint*,d : Database, fk : ForeignKey
\end{align*}

@@

\paragraph{Mapping of Classes} The class is mapped into a table with  automatically created primary key. The primary key is always created with name Id and of the $\mathbb{N^{*}}$ type. The properties and associations has to be mapped separately.
\begin{align*}
	\mathbf{\rho_{c}} : Class*  \rightarrow Table* \\
	\rho_{c}(c) = \begin{cases}
		%\inference{c = \varnothing}{\varnothing}\\\\
		%\inference{|c| > 1}{\rho_c(head(c)) \cup \rho_c(tail(c))} \\ \\
		\inference{|c| = 1}{table(name(c), primaryKey("Id"), \varnothing, \varnothing)}  \end{cases}   \\ 	
 c : Class
\end{align*}

@@

%\paragraph{Get Property or Association} Functions $findProperty$ and $findAssociation$ returns a property (association) if it exists in given context - application and class of provided name.
%\begin{align*}
%	\mathbf{findProperty} : Label \times Label \times Application \rightarrow Property \\
%	findProperty(l_c, l_p, a) = \inference{\exists c \in classes(a): name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_p}{p} \\
%	l_c, l_p \in Label, a : Application
%\end{align*}
%\begin{align*}
%	\mathbf{findAssociation} : Label \times Label \times Application \rightarrow Association \\
%	findAssociation(l_c, l_a, a) = \inference{\exists c \in classes(a): name(c) = l_c \\ \wedge \exists o \in associations(c) : name(o) = l_a}{o} \\
%	l_c, l_a \in Label, a : Application
%\end{align*}

\paragraph{Add Property} The name collision only has to be verified, before adding a property into an application structure:
\begin{align*}
 	\mathbf{addProp} : Label \times Property \times Application \rightarrow Application \\
	addProp(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(c) : name(q) \neq name(p) \\ \wedge \forall e \in associations(c) : name(e) \neq name(p)}{properties(c) \cup p } \\
	l \in Label, p : Property, a : Applications
\end{align*}

\paragraph{Add Association} Adding an association into an application structure has to verify if there is no name collision and if there is the associated class present in the application model.
\begin{align*}
	\mathbf{addAssoc} : Label \times Association \times Application \rightarrow Application \\
	addAssoc(l, s, a) = \inference{
	\exists c \in classes(a) : name(c) = l \\
	\wedge \forall f \in associations(c) : name(f) \neq name(s) \\ 	\wedge \forall p \in properties(c) : name(p) \neq name(s) \\ 
	\wedge \exists k \in classes(a) : name(k) = reference(s)}{associations(c) \cup s}\\
	l \in Labels, s \in Associations, a : Applications
\end{align*}
